---
title: "NHANES data summary for IndxArea SES analysis"
output: html_document
date: "2026-08-10"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,       # Show code in HTML output
  warning = FALSE,   # Hide warning messages
  message = FALSE,    # (Optional) Hide package loading / diagnostic messages
  fig.width=10, fig.height = 4,
  root.dir = "~/Documents/GitHub/nhanes_area_ses/run"
)

## load pre-built pantry functions (for data wrangling)
get_github_scripts <-function(user, repo, path) {
  api <- sprintf("https://api.github.com/repos/%s/%s/contents/%s", user, repo, path)
  scripts <- grep("*.R", jsonlite::fromJSON(api)$name, value = T)
  URLs <- lapply(scripts, function(f) sprintf("https://raw.githubusercontent.com/%s/%s/main/%s/%s", user, repo, path, f))
  invisible(lapply(URLs, source))
} ; get_github_scripts("julieeg", "pantry", "functions")

library(ggpubr) ; library(knitr) ; library(data.table) ; library(ComplexHeatmap) ; 
library(circlize)

###############################################################
## Load pre-built NHANES index tables & variable summaries
###############################################################

## List of all NHANES variables --------------------
nhanes_vars_all <- fread("../data/raw/nhanes_varlist_all.csv") 

## Load project data dictionary to grab years, tables & variables of interest
project_datadict <- readxl::read_xlsx("../run/nhanes_pullrequest_IndVsArea_09012026.xlsx")

# Load processed nhanes data file
nhanes_processed <- readRDS("../data/processed/nhanes_processed.rda") 


#############################################
## Build basic ggplot theme
#############################################

ggplot_theme <- theme_bw() + 
  theme(axis.text = element_text(color="black"),
        panel.grid.minor = element_blank(),
        axis.title = element_text(face="bold"),
        strip.background = element_blank(),
        strip.text.x.top = element_text(face="bold"))
```


### Check variable harmonization across NHANES cycles

First, check how many participants (Age >18 years) are available at each exam cycle
with BOTH the interview & MEC components 

(**Participants with interview only will NOT have laboratory or exam data**)

```{r, check_participant_types}
nhanes_participant_types <- nhanes_processed %>% 
  group_by(Years) %>% 
  reframe(n=n(),
          n_type=table(RIDSTATR)) %>%
  mutate(pct=(n_type/n)*100, 
         Participant_Type = rep(c("MEC_and_Interview", "Interview_Only"), 12)) %>%
  mutate_at("Years", ~factor(., levels=c(rev(unique(nhanes_processed$Years)))))

ggarrange(plotlist=list(
  nhanes_participant_types %>% 
    ggplot(aes(y=Years, group=Participant_Type, x=n_type, fill=Participant_Type)) + 
    geom_bar(stat="identity") + xlab("Distribution (n)") + 
    ggplot_theme,
  nhanes_participant_types %>% 
    ggplot(aes(y=Years, group=Participant_Type, x=pct, fill=Participant_Type)) + 
    geom_bar(stat="identity") + xlab("Distribution (%)") + 
    ggplot_theme
  ), nrow=1, align="hv", common.legend=T)

```

Next, summarise missingness of variables from each data category (DEMO, LAB, EXAM, QUESTIONNAIRE, DIET)
among participants who completed BOTH the Interview and MEC exam

```{r, check_miss_1, fig.height=6}
# select list of variables representing all categories to check missingness
vars_checkmiss <- c("age", "racethn", "smoke_current", "bmi", "sbp_mean", "tc", 
                    "hdl", "uacr", "hba1c", "fg","tg", "egfr", "cvd",
                    "ascvd", "ckd", "diabetes", "rx_use_statin", "rx_use_bpmed",
                    "HEI2015_ALL")

# summarise missingness -- before and after accounting for participant type
na_summary <- nhanes_processed %>%
  select(SEQN, Years, RIDSTATR, vars_checkmiss) %>%
  mutate(SEQN_TYPE = ifelse(RIDSTATR == "Both interviewed and MEC examined", 2, 1)) %>%
  pivot_longer(
    cols = vars_checkmiss, names_to = "variable", values_to = "value",
    values_transform = list(value = as.character)) %>%
  group_by(Years, variable) %>%
  summarize(
    n_all = n(), n_na =  sum(is.na(value)),
    pct_na = round(mean(is.na(value))*100,1),
    n_both = sum(ifelse(SEQN_TYPE == 2,1,0)),
    n_na_both = sum(ifelse(is.na(value) & SEQN_TYPE == 2,1,0)),
    pct_na_both = round((n_na_both/n_both)*100,1),
    .groups    = "drop")

# matrix of missingness, restricting to participants with BOTH data types (for analysis)
na_summary.hm <- na_summary %>%
    select(Years, variable, pct_na_both) %>%
    pivot_wider(names_from=Years, values_from=pct_na_both) %>%
    column_to_rownames("variable") %>%
    as.matrix()

# View Heatmap -----------------
Heatmap(
  na_summary.hm, cluster_columns = F, rect_gp = gpar(col = "white", lwd = 1),
  show_row_dend = F, row_names_side = "left", column_title_side = "bottom",
  column_title = "NHANES Exam Cycles", row_title = "NHANES Variables",
  name = "Missing\nData (%)", row_names_gp = gpar(fontsize=9),
  col=colorRamp2(c(0, 100), c("#FFFFFF", palettes$NatExt$Blues[3])), 
  # Function to print missing value if >10%
  cell_fun = function(j, i, x, y, width, height, fill) {
    pct_na <- as.numeric(na_summary.hm[i, j])
    if (!is.na(pct_na) & pct_na>1) {
      grid.text(sprintf("%.1f", pct_na), x, y, gp = gpar(fontsize = 8, col = "black"))
      }
    })

```
*Note:* Values are provided in the boxes when % missing > 1. 
Missingness would seem much higher if we did not account for participants that only completed the interviews:

```{r, check_miss_2, echo=F}
na_summary_all.hm <- na_summary %>% 
  select(Years, variable, pct_na) %>%
  filter(variable %in% c("tg", "fg", "egfr", "hdl", "tc", "hba1c", "sbp_mean", "smoke_current")) %>%
  pivot_wider(names_from=Years, values_from=pct_na) %>%
  column_to_rownames("variable") %>%
  as.matrix()

Heatmap(
  na_summary_all.hm, cluster_columns = F, rect_gp = gpar(col = "white", lwd = 1),
  show_row_dend = F, row_names_side = "left", column_title_side = "bottom",
  column_title = "NHANES Exam Cycles", row_title = "NHANES Variables",
  name = "Missing\nData (%)", row_names_gp = gpar(fontsize=9),
  col=colorRamp2(c(0, 100), c("#FFFFFF", palettes$NatExt$Teals[3])), 
  # Function to print missing value if >10%
  cell_fun = function(j, i, x, y, width, height, fill) {
    pct_na <- as.numeric(na_summary_all.hm[i, j])
    if (!is.na(pct_na) & pct_na>1) {
      grid.text(sprintf("%.1f", pct_na), x, y, gp = gpar(fontsize = 7, col = "black"))
      }
    })
```

\n
## Descriptive Tables & Figures
### BEFORE applying survey weights

##### Summarise strata distributions, across NHANES cycles

```{r, summarise_descript_1}
# subset participants to those aged >= 18 years with BOTH data types
nhanes1 <- nhanes_processed %>% filter(RIDSTATR != "Interviewed only")

# Check age, sex & race distributions by exam cycle
vars_strata <- c(age_gt65 = "Age <65 years", female="Female", 
                 racethn_addNHA = "Race/Ethnicity (add NHA)")

# View table --------------
print_summary_table(var_strata = "Years", vars_to_summarize = vars_strata, 
                    p_adjust = "none", p_print = F, data = nhanes1) %>% kable()
```

\n
#### Summarise disease outcomes across NHANES cycles:

* Include adults ages >30 and < 80
* Exclude particiants with any CVD, ASCVD, CKD or Diabetes (un/diagnosed)

```{r}
vars_diseases <- c(ascvd="ASCVD", cvd="CVD", ckd="CKD", diabetes="Diabetes", 
                   diabetes_undx="Diabetes (undiagnosed)")
print_summary_table(var_strata = "Years", vars_to_summarize = vars_diseases, 
                    p_adjust = "none", p_print = F, data = nhanes1) %>% kable()
```


##### Examine clincial characteristics between strata, among elligible particiapnts

```{r, summarise_descript_2}

# List clinical variables to check for quality/missingness --------
vars_clin_cont <- c(age="Age, years", bmi="BMI, kg/m2", sbp_mean="SBP (average), mmHg", 
                    dbp_mean="DBP (average), mmHg", tc="Total cholesterol, mg/dL", 
                    hdl="HDL-c, mg/dL", ldl="LDL-c, mg/dL", tg="Triglyeride, mg/dL",
                    hba1c="HbA1c (%)", uacr="UACR", egfr="eGFR", egfr_race = "eGFR (with race)")

# Tabulate table1s by each STRATA -----------
tab1_clin_byASR.l<-lapply(list("age_gt65.lab", "gender", "racethn_addNHA"), function(strata) {
  print_summary_table(data = nhanes1, vars_to_summarize = vars_clin_cont, 
                      var_strata = strata, p_print = F, p_adjust = "none", digits = c(1,1),
                      var_strata_order = as.vector(unique(nhanes1 %>% pull(strata))))
}) ; names(tab1_clin_byASR.l) <- c("Age", "Sex", "Race")


# Build table1 by age, sex & race --------
tab1_clin_byASR <- do.call(cbind.data.frame, tab1_clin_byASR.l) %>% 
  rename(Total = Age.Total) %>%
  select(-ends_with(".Total")) 

## View Table1 by Age & Sex ----------------------
tab1_clin_byASR %>% select(Total, starts_with("Age"), starts_with("Sex")) %>% 
  kable(caption = "Select clinical characteristics of inlucded participants by Age & Sex")

# View Table1 by Race (including NHA) ----------------------
tab1_clin_byASR %>% select(starts_with("Race")) %>% 
  kable(caption = "Select clinical characteristics of inlucded participants by Race/Ethnicity")

## Summarise medication use
tab1_clin_byMED.l<-lapply(list("rx_use_statin", "rx_use_bpmed", "rx_use_diabmed"), function(strata) {
  print_summary_table(data = nhanes1, vars_to_summarize = vars_clin_cont, digits = c(1,1),
                      var_strata = strata, p_print = F, p_adjust = "none")
  }) ; names(tab1_clin_byMED.l) <- c("Statin", "BP med", "Diabetes med")

# View table by Medication use -----------------------------------
do.call(cbind.data.frame, tab1_clin_byMED.l) %>% 
  select(-ends_with("Total")) %>%
  kable(caption = "Select clinical characteristics of inlucded participants by use of Statin/BP/Diabetes medications")

```


### Examine histograms of continuous lab values

Include ALL participants with Age > 18 years

```{r, check_hist_all}
# continuous variables to examine
lab_vals <- c( BMI="bmi", WHR="whr", SBP="sbp_mean",
               DBP="dbp_mean", HbA1c="hba1c", Glucose="glu", Total.Cholesterol="tc",
               HDLc="hdl", LDLc="ldl", Triglyceride="tg", uAlbumin="u_albumin", 
               uCreatinine="u_creatinine", UACR="uacr", eGFR="egfr", eGFR.race = "egfr_race")

# Make long-format data for panel histograms
nhanes1_long <- nhanes1 %>% 
  select(SEQN, lab_vals) %>% 
  pivot_longer(cols=-SEQN) %>%
  mutate_at("name", ~factor(., levels=names(lab_vals)))

# Build generic histogram function
plot_hist.fun <- function(vars, data = nhanes1_long) {
  data %>%
    filter(name %in% vars) %>%
    ggplot(aes(x=value)) + facet_wrap(~name, scale="free", ncol=4) + 
    geom_histogram() + labs(x="Laboratory value", y="Distribution (n)") +
    ggplot_theme
}

## View histograms ----------------------
# Anthropometric & BP vars: BMI, Waist-Hip ratio, SBP, DBP
plot_hist.fun(names(lab_vals[1:4]))
# Blood Glucose: HbA1c and plasma glucose
plot_hist.fun(names(lab_vals[5:6]))
# Blood lipids: HbA1c and plasma glucose
plot_hist.fun(names(lab_vals[7:10]))
# Kidney outcomes: urine albumin, creatinine, UACR and eGFR
plot_hist.fun(names(lab_vals[11:14]))    

```

### Examine missingness & distributions of PREVENT inputs 
```{r}
nhanes_processed_prevent <- readRDS("../data/processed/nhanes_processed_prevent.rda")

# summarise missingness 
prevent_na_summary.hm <- nhanes_processed_prevent %>%
  select(SEQN, Years, starts_with("prevent") & !contains("0yr") & !contains("complete")) %>%
  pivot_longer(
    cols = starts_with("prevent"), names_to = "var", values_to = "value",
    values_transform = list(value = as.character)) %>%
  group_by(Years, var) %>%
  summarize(pct_na = round(mean(is.na(value))*100,1), .groups    = "drop") %>%
  select(Years, var, pct_na) %>%
  pivot_wider(names_from=Years, values_from=pct_na) %>%
  column_to_rownames("var") %>% as.matrix()

# View Heatmap -----------------
Heatmap(
  prevent_na_summary.hm, cluster_columns = F, rect_gp = gpar(col = "white", lwd = 1),
  show_row_dend = F, row_names_side = "left", column_title_side = "bottom",
  column_title = "NHANES Exam Cycles", row_title = "PREVENT Inputs",
  name = "Missing\nData (%)", row_names_gp = gpar(fontsize=9),
  col=colorRamp2(c(0, 100), c("#FFFFFF", palettes$NatExt$Blues[3])), 
  # Function to print missing value if >10%
  cell_fun = function(j, i, x, y, width, height, fill) {
    pct_na <- as.numeric(prevent_na_summary.hm[i, j])
    if (!is.na(pct_na) & pct_na>0) {
      grid.text(sprintf("%.1f", pct_na), x, y, gp = gpar(fontsize = 8, col = "black"))
      }
    })
```


##### Summarise characteristics of participants for PREVENT analyses:

 * Adults 30 to 79 years of age 
 * Without known ASCVD, CVD, CKD or diabetes 
 * **TENTATIVE: No extreme clinical ranges for SBP, TC, HDL-C (SBP <90 or >200 mm Hg, TC <130 or >320 mg/dL, and HDL-C <20 or >100 mg/dL).**
 * **TENTATIVE: No extreme values for BMI (<18.5 or ≥40.0 kg/m2)**

```{r, check_hist_include}

# Merge clinical/lab outcomes into PREVENT dataset, for descriptive analyses
nhanes_processed_prevent <- left_join(
  nhanes_processed_prevent, 
  nhanes_processed %>% select(SEQN, age_gt65.lab, gender, bmi, 
                              tc, tc, hdl, tg, hba1c, uacr, egfr, egfr_race),
  by="SEQN"
)

vars_prevent <- c(prevent_age="Age, years", prevent_sbp = "SBP, 90-180 mmHg", 
                  prevent_bprx = "Use BP medications", prevent_tc = "TC, 130-320 mg/dL", 
                  prevent_hdl = "HDLc, 20-100", prevent_statin = "Use statins", 
                  prevent_diab = "Has diabetes", prevent_smoking = "Current smoker", 
                  prevent_egfr = "eGFR, 15-140", prevent_egfr_race = "eGFR (race), 15-140", 
                  prevent_bmi = "BMI, 18.5-40", prevent_hba1c = "HbA1c, 4.5-15 %", 
                  prevent_uacr = "UACR, 0.1-25000")

tab1_prevent_byASR.l <- lapply(list("age_gt65.lab", "gender", "racethn_addNHA"), function(strata) {
  print_summary_table(data = nhanes_processed_prevent, vars_to_summarize = vars_prevent, 
                      var_strata = strata, p_print = F, p_adjust = "none", digits = c(1,1),
                      var_strata_order = as.vector(unique(nhanes1 %>% pull(strata))))
}) ; names(tab1_prevent_byASR.l) <- c("Age", "Sex", "Race")


# Build table1 by age, sex & race --------
tab1_prevent_byASR <- do.call(cbind.data.frame, tab1_prevent_byASR.l) %>% 
  rename(Total = Age.Total) %>%
  select(-ends_with(".Total")) 

## View Table1 by Age & Sex ----------------------
tab1_prevent_byASR %>% select(Total, starts_with("Age"), starts_with("Sex")) %>% 
  kable(caption = "PREVENT inputs among participants by Age & Sex")

# View Table1 by Race (including NHA) ----------------------
tab1_prevent_byASR %>% select(starts_with("Race")) %>% 
  kable(caption = "PREVENT inputs among participants by Race/Ethnicity")

```



